---
title: "internal validation - unseen species predicted from phage bank"
author: "MC, FF"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(janitor) #rounding up at 0.5
library(tidyverse)
```

```{r,echo=FALSE}
load("intval_n100_train4_2024.RData")
```

## General properties

 * We are interested in prediciting the number of unseen species given a data set of already sampled individuals
 * Based on two dumps of the phage database: September 2024 and May 2025
 * We have performed 100 validation runs per host species, per year and per training data set size, where we divided the whole sampled species vector, per host species, *randomly* into training and validation sets with a predetermined size of the training set (in %, rounded up, of the full data set)
 * We assessed the four estimators Good-Toulmin [GT], Efron-Thisted [ET], Empirical Bayes approach using an underlying Pitman-Yor-Process model (2 parameter) [PYP] and the (effectively one-dimensional) parametric Fisher-Poisson-Gamma model [FPG] for predicting the number of new (unseen) species in the validation subset   
 * Error measure NMAE (normalized mean absolute error): records the absolute error between prediction and true value divided by the true value (=in units of the true value)
    * we report the median of the NMAE across the 100 simulations [**We can add 25% and 75% without rerunning, I recorded summaries across the 100 sims in each parameter combination**]
 

## Data from September 2024 - median NMAE


```{r}
res1 <- sapply(valid_res,
               function(x){signif_half_up(x["NMAE: Median",],2)})
#colnames(res1) <- abbreviate(colnames(res1),minlength = 6)
kable(res1)
```



\newpage

## Data from May 2025 - median NMAE

```{r}
load("intval_n100_train4_2025.RData")
#remove bajr
valid_res <- valid_res[-9]
res2 <- sapply(valid_res,
               function(x){signif_half_up(x["NMAE: Median",],2)})
colnames(res2) <- abbreviate(colnames(res2),minlength = 6)
kable(res2)
load("intval_n100_train4_2025.RData")
#remove bajr
valid_res <- valid_res[-9]
res2 <- sapply(valid_res,
               function(x){signif_half_up(x["NMAE: Median",],2)})
```

\newpage

```{r plot_prep_24}
plotd <- res1
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> separate_wider_delim(cols = stat,delim = ".0.",names = c("stat","training"))
plotd <- plotd |>  mutate(stat = fct_relevel(stat, 
            c("GT","ET","PYP","FPG")))
plotd2024 <- plotd
```

```{r plot_prep_24}
plotd <- res2
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> separate_wider_delim(cols = stat,delim = ".0.",names = c("stat","training"))
plotd <- plotd |>  mutate(stat = fct_relevel(stat, 
            c("GT","ET","PYP","FPG")))
plotd2025 <- plotd
```


## Plots

```{r}
ggplot(data = plotd2024,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2024) + facet_wrap(vars(training),scales="free_y")
```

```{r}
ggplot(data = plotd2025,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2025) + facet_wrap(vars(training),scales="free_y")
```


```{r}
ggplot(data = plotd2024,aes(x=training,y=value,colour = stat,group = stat)) +   geom_point() +   geom_line() + labs(title = 2024) + lims(y=c(0,1)) + facet_wrap(vars(Host),scales="free_y")
```
```{r}
ggplot(data = plotd2025,aes(x=training,y=value,colour = stat,group = stat)) +   geom_point() +   geom_line() + labs(title = 2025) + lims(y=c(0,0.2)) + facet_wrap(vars(Host),scales="free_y")
```

 * There are issues with the FPG estimator (2 missing data points)
 * As expected, the GT estimator breaks down when predicting larger samples from smaller samples. The value is capped at +2, but actually is much larger for some species

```{r}
# 2024
plotd <- res1
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.25,ET.0.25,PYP.0.25,FPG.0.25)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.25","ET.0.25","PYP.0.25","FPG.0.25")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2024)
#2025
plotd <- res2
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.25,ET.0.25,PYP.0.25,FPG.0.25)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.25","ET.0.25","PYP.0.25","FPG.0.25")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2025)

```

## Plots - median NMAE, 2024/25, training on 50%

 * There are issues with the FPG estimator (missing data point)

```{r}
# 2024
plotd <- res1
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.5,ET.0.5,PYP.0.5,FPG.0.5)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.5","ET.0.5","PYP.0.5","FPG.0.5")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2024)
#2025
plotd <- res2
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.5,ET.0.5,PYP.0.5,FPG.0.5)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.5","ET.0.5","PYP.0.5","FPG.0.5")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2025)

```

## Plots - median NMAE, 2024/25, training on 65%

```{r}
# 2024
plotd <- res1
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.65,ET.0.65,PYP.0.65,FPG.0.65)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.65","ET.0.65","PYP.0.65","FPG.0.65")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2024)
#2025
plotd <- res2
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.65,ET.0.65,PYP.0.65,FPG.0.65)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.65","ET.0.65","PYP.0.65","FPG.0.65")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2025)

```

## Plots - median NMAE, 2024/25, training on 80%

```{r}
# 2024
plotd <- res1
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.8,ET.0.8,PYP.0.8,FPG.0.8)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.8","ET.0.8","PYP.0.8","FPG.0.8")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2024)
#2025
plotd <- res2
plotd <- apply(plotd,c(1,2),function(x){min(x,2)})
plotd <-  data.frame(t(plotd))|> rownames_to_column(var = "Host") 
plotd <- plotd |> select(Host,GT.0.8,ET.0.8,PYP.0.8,FPG.0.8)
plotd <- plotd |> pivot_longer(cols = !Host,names_to = "stat",
                               values_to = "value")
plotd <- plotd |> mutate(stat = fct_relevel(stat, 
            c("GT.0.8","ET.0.8","PYP.0.8","FPG.0.8")))
ggplot(data = plotd,aes(x=stat,y=value,colour = Host,group = Host)) + 
  geom_point() +   geom_line() + labs(title = 2025)

```


